
---
title: "COVID_FINAL"
author: "Chris Kaldrovics"
date: "2025-03-24"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(tidyverse)
library(dplyr)
library(ggplot2)
```

## Data Import and Clean Up

This is where I loaded in my Data and Cleaned it: 
To do so I loaded in via URL and file names. I then concatenated the two and read in the csv files from those links to create some tables:

global_cases
global_deaths
US_cases
US_deaths

```{r,echo=FALSE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv")

urls <- str_c(url_in,file_names)

US_cases <- read_csv(urls[1])
global_cases <-read_csv(urls[2])
US_deaths <-read_csv(urls[3])
global_deaths<-read_csv(urls[4])
```

After Loading My data in I cleaned it and trimmed it down for Analysis Ultimately creating a table called global where I will delve deeper into the data. 
I created a table called global and then looked at the summary of that table to define changes necessary for analysis

```{r,echo=TRUE}
global_cases <- global_cases %>%
     pivot_longer(cols = -c('Province/State',
                            'Country/Region',Lat,Long),
                  names_to = "date",
                  values_to = "cases") %>%
     select(-c(Lat,Long))
global_deaths <- global_deaths %>%
     pivot_longer(cols = -c('Province/State',
                            'Country/Region',Lat,Long),
                  names_to = "date",
                  values_to = "deaths") %>%
     select(-c(Lat,Long))
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country/Region',
          Province_State = 'Province/State') %>%
  mutate(date = mdy(date))
 

summary(global)
```
After that summary I decided it is best to filter out rows that have 0 cases and check to ensure the rest of the data is accurate. 
```{r,echo=FALSE}
global<-global %>% filter(cases > 0)
global %>% filter(cases > 28000000)
```
# Analysis and Visualizations
Now my data is ready for some visuals and a model! 

## COVID_19 DEATHS 
```{r, echo=FALSE}
# Find the top 5 countries based on total deaths
top_deaths_countries <- global %>%
  group_by(Country_Region) %>%
  summarise(Max_Deaths = max(deaths), .groups = 'drop') %>%
  top_n(5, Max_Deaths)
# Plotting the bar chart
ggplot(top_deaths_countries, aes(x = reorder(Country_Region, Max_Deaths), y = Max_Deaths, fill = Country_Region)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 5 Countries by Total COVID-19 Deaths",
       x = "Country/Region",
       y = "Maximum Deaths") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

The above plot tells us that covid had the biggest impact in these countries especially the United States. 

# The US Covid Cases Over Time
How many people got sick in the United States?
```{r,echo=FALSE}


# Assuming your data frame is named 'global'
us_cases_over_time <- global %>%
  filter(Country_Region == "US") %>%
  arrange(date) %>%
  select(date, cases)  # Assuming 'cases' is the cumulative cases field

# View the data structure
head(us_cases_over_time)
ggplot(us_cases_over_time, aes(x = as.Date(date), y = cases)) +
  geom_line(color = "blue") +
  labs(title = "Cumulative COVID-19 Cases in the US Over Time",
       x = "Date",
       y = "Cumulative Cases") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

The above chart shows us COVID-19 cases cumulatively in the United States.

It is crazy to believe that in just a few years a virus can get 100,000,000 people sick.

Some more cleaning for our model! 
```{r,echo=FALSE}
# Adjust the pivot_longer() function
US_deaths_long <- US_deaths %>%
     pivot_longer(
         cols = -(UID:Population),  # This uses range exclusion based on your column list
         names_to = "date",
         values_to = "deaths"
     )
```
# Model
Do COVID 19 Deaths have to do with State Population in the United States?
```{r,echo=TRUE}
# Calculate the correlation coefficient
correlation_coefficient <- cor(US_deaths_long$Population, US_deaths_long$deaths, use = "complete.obs")
 correlation_coefficient

# Perform the correlation test
correlation_test <- cor.test(US_deaths_long$Population, US_deaths_long$deaths, method = "pearson", use = "complete.obs")

# Print the results
print(correlation_test)

```
Yes Population in certain states is Positively Correlated to COVID Deaths!
Thanks!
